---
title: "CoxPH"
author: Calvin Belton & James Walters
format: html
editor: source
theme: sandstone
toc: true
embed-resources: true
number-sections: true
---

# Setting up

```{r}
library(flexsurv)
library(ggplot2)
library(tidyverse)
library(ggfortify)# for `autoplot` with survfit
library(survminer)
library(kableExtra)
library(patchwork)
library(splines)
library(muhaz)
library(reporter)
library(gtsummary)
```

## Load previously 'cleaned' dataset

```{r}
load(file = "../Input_Data/monsurv.Rdata")
```

## Function for saving files

```{r}
save_file <- function(test_obj, label, file){

  cat(
    "\n-----------------------------------\n",
    label,
    "\n-----------------------------------\n",
    file = file,
    append = TRUE
  )

  capture.output(
    test_obj,
    file = file,
    append = TRUE
  )
}
```

```{r}
survival_stats <- "../Results/Survival_CoxPH/Statistics/survival_stats.txt"

cat(
  "Monarch Manuscript Survival Results\n",
  "=================================\n\n",
  file = survival_stats
)
```

```{r}
mortality_stats <- "../Results/Survival_CoxPH/Statistics/mortality_statistics.txt"

cat(
  "Monarch Manuscript Mortality Results\n",
  "=================================\n\n",
  file = mortality_stats
)
```

```{r}
coxph_stats <- "../Results/Survival_CoxPH/Statistics/coxph_statistics.txt"

cat(
  "Monarch Manuscript Cox Proportional Hazards Results\n",
  "=================================\n\n",
  file = coxph_stats
)
```

```{r}
anova_stats <- "../Results/Survival_CoxPH/Statistics/ANOVA_statistics.txt"

cat(
  "Monarch Manuscript Cox Proportional Hazards ANOVA Results\n",
  "=================================\n\n",
  file = anova_stats
)
```

```{r}
diagnostics_stats <- "../Results/Survival_CoxPH/Statistics/diagnostics_statistics.txt"

cat(
  "Monarch Manuscript Diagnostics of Cox Proportional Hazards Models\n",
  "=================================\n\n",
  file = diagnostics_stats
)
```

## Graphics

Before I start creating plots, I set up the custom theme and custom color palette that is continuous in my plots of monarch. 
## Custom theme

```{r}
theme_monarch <- theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(face = "bold", size = 13),
    axis.text = element_text(face = "bold", size = 11),
    strip.text = element_text(face = "bold")
  )
```

## Graph systematics

```{r}
sex_colors <- c("Male" = "dodgerblue","Female" = "firebrick2")
```

# Sex Comparison of Survival

## Create Survival Object

```{r}
survMon <- Surv(time = monsurv$Age, event = monsurv$event)
```

## Survdiff created: Kaplan-Mier statistics

```{r}
survMon.diff <- survdiff(survMon ~ Sex, data = monsurv)
print(survMon.diff)
save_file(survMon.diff,"Kaplan-Mier statistics", survival_stats)
```

### Find the p-value

```{r}
p <- survMon.diff$pvalue
save_file(p,"Kaplan-Mier p-value", survival_stats)
```

### Analyze with survfit

```{r}
monsurv.fit <- survfit(Surv(Age) ~ Sex, data = monsurv)
print(monsurv.fit)
save_file(monsurv.fit,"Kaplan-Mier survfit survival estimation", survival_stats)
```

## Kaplan-Meier Plot

```{r}
survivorship_plot <- ggsurvplot(
  monsurv.fit,
  conf.int = FALSE,
  palette = sex_colors,
  xlab = "Days Since Eclosion",
  ylab = "Survival Probability",
  legend.title = "Sex",
  legend.labs = c("Male", "Female"),
  legend = "right",
  size = 1,
  ggtheme = theme_monarch
)
print(survivorship_plot)
```

### Save Kaplan-Meier Plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_01_survivorship_plot.pdf",
  plot = survivorship_plot$plot,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

# Mortality rate visualization

## Calculate qx ()

### Summarize previously created survfit

```{r}
mon_sum <- summary(monsurv.fit)
str(mon_sum)
```

### Generate life table, including generating qx

```{r}
mon_sum_DF <- data.frame(
  x   = mon_sum$time,
  nx = mon_sum$n.risk,
  lx   = mon_sum$surv,
  dx = mon_sum$n.event,
  lower  = mon_sum$lower,
  upper  = mon_sum$upper,
  strata = sub(x = mon_sum$strata, pattern = "Sex=", replacement = "")
)
mon_sum_DF$qx <- mon_sum_DF$dx/mon_sum_DF$nx
save_file(mon_sum_DF,"Life-table data", mortality_stats)
```

### Visualize qx and log(qx) by sex

```{r}
qxplot <- ggplot(mon_sum_DF, aes(x = x, y = qx, color = strata)) +
  geom_line() +
  geom_point(size = 1) +
  scale_color_manual(values = sex_colors) +
  labs(title = "Mortality over time by Sex",
       x = NULL) +
  theme_bw()

qxLogplot <- ggplot(mon_sum_DF, aes(x = x, y = log(qx), color = strata)) +
  geom_line() +
  geom_point(size = 1) +
  scale_color_manual(values = sex_colors) +
  labs(x = "Days") +
  theme_bw()

mortality_plot <- (qxplot / qxLogplot)
mortality_plot
```

### Save mortality plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_02_mortality_plot.jpg",
  plot = mortality_plot,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

## Generate smoothed hazard rates via muhaz by sex

Use the muhaz package to get kernal-density smoothing of hazard rates. We will only estimate hazard rates up to day 45, because otherwise there is not sufficient data (e.g. remaining individuals) to estimate values with acceptable accuracy.

```{r}
# split by sex
haz_female <- muhaz(
  times = monsurv$Age[monsurv$Sex == "Female"],
  delta = monsurv$event[monsurv$Sex == "Female"],
  max.time = 45
)

haz_male <- muhaz(
  times = monsurv$Age[monsurv$Sex == "Male"],
  delta = monsurv$event[monsurv$Sex == "Male"],
  max.time = 45
)

haz_df <- rbind(
  data.frame(
    age = haz_female$est.grid,
    hazard = haz_female$haz.est,
    loghaz = log(haz_female$haz.est),
    Sex = "Female"
  ),
  data.frame(
    age = haz_male$est.grid,
    hazard = haz_male$haz.est,
    loghaz = log(haz_male$haz.est),
    Sex = "Male"
  )
)
```

### Visualize muhaz by sex
```{r}
muhaz <- ggplot(haz_df, aes(x = age, y = hazard, color = Sex)) +
  geom_line() +
  scale_color_manual(values = sex_colors) +
  coord_cartesian(ylim = c(0, 0.25)) +
  labs(x = "Days", y = "Hazard Rate") +
  theme_bw()

log_muhaz <- ggplot(haz_df, aes(x = age, y = loghaz, color = Sex)) +
  geom_line() +
  scale_color_manual(values = sex_colors) +
  coord_cartesian(ylim = c(-8, 0)) +
  labs(x = "Days", y = "Log(Hazard Rate)") +
  theme_bw()
full_muhaz <- muhaz / log_muhaz
full_muhaz
```
### Save smoothed muhaz plot
```{r}
ggsave(
  "../Results/Survival_CoxPH/Figures/Figure_03_smoothed_muhaz_hazard_plots.jpg",
  plot = full_muhaz,
  width = 180,
  height = 120,
  units = "mm",
  dpi = 300
)
```


### Plot overlaying manually calculated qx with the smoothed estimates.
```{r}
liftab.sex <- split(mon_sum_DF, mon_sum_DF$strata)

jpeg("../Results/Survival_CoxPH/Figures/Figure_04_muhaz_overlaid_by_qx.jpg", width = 2400, height = 1200, res = 300)

par(
  mfrow = c(1,2),
  mar = c(4, 4, 2, 1) 
)

plot(log(haz_female$haz.est) ~ haz_female$est.grid, 
     col = sex_colors["Female"], type = "l", ylim = c(-8,0), xlim = c(0,50),
     ylab = "Log(Hazard Rate)", xlab = "Days", main = "Female")
points(log(liftab.sex$Female$qx) ~ liftab.sex$Female$x, 
       col = sex_colors["Female"])

plot(log(haz_male$haz.est) ~ haz_male$est.grid, 
     col = sex_colors["Male"], type = "l", ylim = c(-8,0), xlim = c(0,50),
     ylab = "Log(Hazard Rate)", xlab = "Days", main = "Male")
points(log(liftab.sex$Male$qx) ~ liftab.sex$Male$x, 
       col = sex_colors["Male"])

dev.off()

```
# CoxPH modeling with wing-length

## Modeling using wing-length

```{r}
m0_nosex <- coxph(Surv(Age, event) ~ WingLength, data = monsurv)
sum_m0_nosex <- summary(m0_nosex)
save_file(sum_m0_nosex,"CoxPH model(WingLength)", coxph_stats)
```

## Model with wing-length and sex

```{r}
m0_sex <- coxph(Surv(Age, event) ~ Sex + WingLength, data = monsurv)
sum_m0_sex <- summary(m0_sex)
save_file(sum_m0_sex,"CoxPH model(WingLength + Sex)", coxph_stats)
```

## ANOVA comparison of the two previous models

```{r}
anova_one <- anova(m0_nosex, m0_sex, test="LRT")
save_file(anova_one,"Anova Comparison CoxPH model(WingLength) vs. CoxPH model(WingLength + Sex)", anova_stats)
```

## Contrast survival between males WITH wing-length vs MISSING wing-length, using KM plot

```{r}
monsurv$male_group <- NA

monsurv$male_group[monsurv$Sex == "Male" & is.na(monsurv$WingLength)] <- "NoWing"
monsurv$male_group[monsurv$Sex == "Male" & !is.na(monsurv$WingLength)] <- "Wing"

male_data <- subset(monsurv, Sex == "Male")

fit <- survfit(Surv(Age, event) ~ male_group, data = male_data)

male_plot <- ggsurvplot(
  fit, data = male_data,
  pval = FALSE,
  risk.table = FALSE,
  legend.title = "Male Group",
  legend.labs = c("No Winglength Measured", "Winglength Measured"),
  legend = c(0.8, 0.5),
  palette = c("forestgreen", "darkorange"),
  ggtheme = theme_monarch,
  xlab = "Days Since Eclosion"
)
print(male_plot)
```

### Save male_plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_05_male_plot.pdf",
  plot = male_plot$plot,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

### Do KM test via survdiff for two groups of males (demonstrates that males missing wings are longer lived)

```{r}
km_test <- survdiff(Surv(Age, event) ~ male_group, data = male_data)
print(km_test)
save_file(km_test,"Kaplan-Mier statistics for males data", survival_stats)
```

# CoxPH with Forward Discal Cell

## Standard fit

Run CoxPH using Sex and normalized size (the z-transformed forward wing discal cell length) as predictors. 
## Model with only size.
```{r}
m0_nosex <- coxph(Surv(Age, event) ~ FWdc_z, data = monsurv)
summary(m0_nosex)
```
## Model with size + sex.
```{r}
m0_sex <- coxph(Surv(Age, event) ~ Sex + FWdc_z, data = monsurv)
summary(m0_sex)
```

## Perform LRT via anova to see if there is a significantly better fit with Sex.This is not the case.
```{r}
anova(m0_nosex, m0_sex, test="LRT")
```

## Diagnostics

Now consider Schoenfeld diagnostics for these predictors. Can check size only, but it is essentially the same with a model with sex included.

```{r}
czNoSex <- cox.zph(m0_nosex)
print(czNoSex)
save_file(czNoSex,"Diagnostics of CoxPH model(Forewing discal cell)", diagnostics_stats)
```

### Plot diagnostics

```{r}
par(mfrow = c(1, 2))

yLim <- c(-5, 5)

plot(
  czNoSex,
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

pval <- czNoSex$table[1, "p"]

text(
  x = 0.5,
  y = 4,
  labels = paste0("p = ", format.pval(pval, digits = 3))
)
```
### Save plot

```{r}
jpeg(
  filename = "../Results/Survival_CoxPH/Figures/Figure_06_size_coxph_diagnostics.jpg",
  width = 180,
  height = 120,
  units = "mm",
  res = 300
)

par(mfrow = c(1, 2))

yLim <- c(-5, 5)

plot(
  czNoSex,
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

pval <- czNoSex$table[1, "p"]

text(
  x = 0.5,
  y = 4,
  labels = paste0("p = ", format.pval(pval, digits = 3))
)

dev.off()
```


## Diagnostics for model with both size and sex.

```{r}
cz0 <- cox.zph(m0_sex)
print(cz0)
save_file(cz0,"Diagnostics of CoxPH model(Forewing discal cell + Size)", diagnostics_stats)
```

### Plot diagnostics

```{r}
par(mfrow = c(1, 2))

yLim <- c(-5, 5)

# Sex
plot(cz0, var = "Sex",
     pch = 19, cex = 0.2,
     col = "red", ylim = yLim)

p_sex <- cz0$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)

# FWdc_z
plot(cz0, var = "FWdc_z",
     pch = 19, cex = 0.2,
     col = "red", ylim = yLim)

p_fw <- cz0$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)
```
### Save plot
```{r}
jpeg(
  filename = "../Results/Survival_CoxPH/Figures/Figure_07_size_+_sex_coxph_diagnostics.jpg",
  width = 180,
  height = 120,
  units = "mm",
  res = 300
)

par(mfrow = c(1, 2))

yLim <- c(-5, 5)

# Sex
plot(cz0, var = "Sex",
     pch = 19, cex = 0.2,
     col = "red", ylim = yLim)

p_sex <- cz0$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)

# FWdc_z
plot(cz0, var = "FWdc_z",
     pch = 19, cex = 0.2,
     col = "red", ylim = yLim)

p_fw <- cz0$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)

dev.off()
```
We have a significant violation of CoxPH with time, where larger individuals have (slightly) lower mortality early in life, but seemingly slightly greater late in life. But the relationship isn’t obviously linear, and perhaps even sinusoidal.

So we need to find some way to account for this time-dependency in the model.

## Start-Stop Split-Time

The Start-Stop Split-Time approach incorporates time-dependency in the model by appropriately capturing the “risk-set” comparisons through manually splitting the data set into many separate, time-limited, observations of survival. Here we create a new data structure were each row corresponds to a short interval of time.

Doing this should give us the best of both worlds: correct management of “risk-set” comparisons while allowing model-testing and diagnostics. 
### Create split

```{r}
cuts <- seq(0, max(monsurv$Age), by = 1)  # 1-day intervals

ds <- survSplit(
  Surv(Age, event) ~ .,
  data = monsurv,
  cut = cuts[-c(1, length(cuts))],  
  start = "tstart",
  end = "tstop",
  event = "event"
)
# define analysis time for the interval (using midpoint)
ds$tmid <- (ds$tstart + ds$tstop) / 2
ds$logtmid <- log(ds$tmid)
```

### Fit split models

```{r}
m_nosex_split <- coxph(Surv(tstart, tstop, event) ~ FWdc_z, data = ds)

m_sex_split <- coxph(Surv(tstart, tstop, event) ~ Sex + FWdc_z, data = ds)

m_nosex_linear_split <-  coxph(Surv(tstart, tstop, event) ~ FWdc_z + FWdc_z:tmid, data = ds)

m_linear_split <-  coxph(Surv(tstart, tstop, event) ~ Sex + FWdc_z + FWdc_z:tmid, data = ds)

m_nosex_log_split <- coxph(Surv(tstart, tstop, event) ~ FWdc_z + FWdc_z:logtmid, data = ds)

m_log_split <- coxph(Surv(tstart, tstop, event) ~ Sex + FWdc_z + FWdc_z:logtmid, data = ds)
```

## size model (ST)
```{r}
m_nosex_split_sum <- summary(m_nosex_split)
print(m_nosex_split_sum)
save_file(m_nosex_split_sum,"CoxPH Model(Forewing discal cell)", coxph_stats)
```
## size + sex model(ST)
```{r}
m_sex_split_sum <- summary(m_sex_split)
print(m_sex_split_sum)
save_file(m_sex_split_sum,"CoxPH Model(Forewing discal cell + Sex)", coxph_stats)
```
## Size + Size:Age model (ST)
```{r}
m_nosex_linear_split_sum <- summary(m_nosex_linear_split)
save_file(m_nosex_linear_split_sum,"CoxPH Model(Forewing discal cell + Forewing discal cell:Age)", coxph_stats)
```

## Sex + Size + Size:Age model (ST)
```{r}
m_linear_split_sum <- summary(m_linear_split)
save_file(m_linear_split_sum,"CoxPH Model(Forewing discal cell + Sex + Forewing discal cell:Age)", coxph_stats)
```

## Size + Size:log(Age) model (ST)
```{r}
m_nosex_log_split_sum <- summary(m_nosex_log_split)
save_file(m_nosex_log_split_sum,"CoxPH Model(Forewing discal cell + Forewing discal cell:logAge)", coxph_stats)
```

## Sex + Size + Size:log(Age) model (ST)
```{r}
m_log_split_sum <- summary(m_log_split)
save_file(m_log_split_sum,"CoxPH Model(Forewing discal cell + Sex + Forewing discal cell:logAge)", coxph_stats)
```

### Split model tests

We can compare nested models with LRT, but the linear and log models have the same number of parameters, and are not nested, so these need to be compared via AIC.

First, consider that we have four possible models:

With or without Sex with or without time-variation in size. So we should do model tests that:

Assess size time-variation without Sex This tests wither time-variation “helps” the model fit, ignoring any role of sex. Assess Sex with size time-variation This tests whether including time-variation impacts the significance of sex as a predictor Assess size time-variation with Sex This tests whether the significance of time-variation depends on the inclusion of sex. We can do these three sets of comparisons for the linear and log temporal relationships. But how to decide between linear vs log, since they have the same number of parameters? We will consider both AIC and model diagnostics.

First, we can simply confirm that the split-time models give the same “simple” result that Sex is not significant without considering time-variation in size.

```{r}
# Confirm absence of "Sex" effect in split-time without time-variation
anova_two <- anova(m_nosex_split, m_sex_split, test = "LRT")
save_file(anova_two,"Anova Comparison CoxPH model(Forewing discal cell) vs. CoxPH model(Forewing discal cell + Sex)", anova_stats)
```

### Compare linear models

```{r}
# Test if linear TV in size improves fit, excluding Sex
anova_three <- anova(m_nosex_split, m_nosex_linear_split,  test = "LRT")
save_file(anova_three,"Anova Comparison CoxPH Model(Forewing discal cell) vs. CoxPH Model(Forewing discal cell + Forewing discal cell:Age)", anova_stats)
```

```{r}
# Compare "Sex" for linear time variation
anova_four <- anova(m_nosex_linear_split, m_linear_split, test = "LRT")
save_file(anova_four,"Anova Comparison CoxPH Model(Forewing discal cell + Forewing discal cell:Age) vs. CoxPH Model(Forewing discal cell + Sex + Forewing discal cell:Age)", anova_stats)
```

```{r}
# Test if linear TV in size improves fit, including Sex
anova_five <- anova(m_sex_split, m_linear_split, test = "LRT")
save_file(anova_five,"Anova Comparison CoxPH model(Forewing discal cell + Sex) vs. CoxPH model(Forewing discal cell + Sex + Forewing discal cell:Age)", anova_stats)
```

So including linear effects of time-variation on size significantly improves model fit, regardless of including Sex. And Sex remains a non-significant predictor even with including linear time-variation for size.

## Compare log models

```{r}
# Test if log TV in size improves fit, excluding Sex
anova_six <- anova(m_nosex_split, m_nosex_log_split,  test = "LRT")
save_file(anova_six,"Anova Comparison CoxPH model(Forewing discal cell) vs. CoxPH model(Forewing discal cell + Forewing discal cell:logAge)", anova_stats)
```

```{r}
# Compare "Sex" for log time variation
anova_seven <- anova(m_nosex_log_split, m_log_split, test = "LRT")
save_file(anova_seven,"Anova Comparison CoxPH model(Forewing discal cell + Forewing discal cell:logAge) vs. CoxPH model(Sex + Forewing discal cell + Forewing discal cell:logAge)", anova_stats)
```

```{r}
# Test if linear TV in size improves fit, including Sex
anova_eight <- anova(m_sex_split, m_log_split, test = "LRT")
save_file(anova_eight,"Anova Comparison CoxPH model(Sex + Forewing discal cell) vs. CoxPH model(Sex + Forewing discal cell + Forewing discal cell:logAge)", anova_stats)
```

Similarly, including log effects of time-variation on size significantly improves model fit, regardless of including Sex. And Sex remains a non-significant predictor even with including log time-variation for size.

## Comparison via AIC

Let’s compute AIC on each of these models, even though the only ones we need this for are linear vs log.

```{r}
aic <- AIC(m_nosex_split, m_sex_split, 
          m_nosex_linear_split, m_linear_split, 
          m_nosex_log_split, m_log_split)

kbl(aic) %>%  kable_classic("striped") %>% 
  scroll_box(width = "60%")
```

Given the negligible AIC difference between linear vs log, linear may be preferable as the “simpler” model. And there is not significantly better fit for the spline model vs either via LRT.

However, we also need to inspect the diagnostics for each of these.

## Split diagnostics log-linear

We can examine the diagnostics for these model fits now. What we find in this case is that Sex now violates the diagnostics for both log and linear, seemingly invoking the mid-life shifts we observed in the plots of interval hazard rates.

However, wingsize also significantly violates model assumptions of constant proportional hazards under the linear model. 
### Diagnosing the linear model

```{r}
linsplit.zph <- cox.zph(m_linear_split)
print(linsplit.zph)
save_file(linsplit.zph,"Diagnostics of CoxPH model(Forewing discal cell + Sex + Forewing discal cell:Age)", diagnostics_stats)
```

### View plot in markdown

```{r linear_coxph_diagnostics, fig.width=8, fig.height=8, dpi=300, out.width="180mm"}

par(mfrow = c(2, 2))

yLim <- c(-5, 5)

# FWdc_z
plot(
  linsplit.zph,
  var = "FWdc_z",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_fw <- linsplit.zph$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)

# Sex
plot(
  linsplit.zph,
  var = "Sex",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_sex <- linsplit.zph$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)

# Size:time
plot(
  linsplit.zph,
  var = "FWdc_z:tmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_size_time <- linsplit.zph$table["FWdc_z:tmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_size_time, digits = 3)),
  cex = 1.1
)
```
### Save plot

```{r linear_coxph_plot}
jpeg(
  filename = "../Results/Survival_CoxPH/Figures/Figure_08_linear_split_coxph_diagnostics.jpg",
  width = 180,
  height = 160,
  units = "mm",
  res = 300
)

par(mfrow = c(2, 2))

yLim <- c(-5, 5)

# FWdc_z
plot(
  linsplit.zph,
  var = "FWdc_z",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_fw <- linsplit.zph$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)

# Sex
plot(
  linsplit.zph,
  var = "Sex",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_sex <- linsplit.zph$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)

# Size:time
plot(
  linsplit.zph,
  var = "FWdc_z:tmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_size_time <- linsplit.zph$table["FWdc_z:tmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_size_time, digits = 3)),
  cex = 1.1
)

dev.off()
```

### Diagnosing the log model

```{r}
logsplit.zph <- cox.zph(m_log_split)
print(logsplit.zph)
save_file(logsplit.zph,"Diagnostics of CoxPH model(Forewing discal cell + Sex + Forewing discal cell:logAge)", diagnostics_stats)
```

### View plot in markdown

```{r log_coxph_diagnostics, fig.width=8, fig.height=8, dpi=300, out.width="180mm"}
par(mfrow = c(2,2))
yLim <- c(-5, 5)

# Sex
plot(
  logsplit.zph,
  var = "Sex",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_sex <- logsplit.zph$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)


# FWdc_z
plot(
  logsplit.zph,
  var = "FWdc_z",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_fw <- logsplit.zph$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)


# FWdc_z:logtmid
plot(
  logsplit.zph,
  var = "FWdc_z:logtmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_log_fw <- logsplit.zph$table["FWdc_z:logtmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_log_fw, digits = 3)),
  cex = 1.1
)
```

### Save plot

```{r log_coxph_plot}
jpeg(
  filename = "../Results/Survival_CoxPH/Figures/Figure_09_log_split_coxph_diagnostics.jpg",
  width = 180,
  height = 160,
  units = "mm",
  res = 300
)

par(mfrow = c(2, 2))
yLim <- c(-5, 5)

# Sex
plot(
  logsplit.zph,
  var = "Sex",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_sex <- logsplit.zph$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)


# FWdc_z
plot(
  logsplit.zph,
  var = "FWdc_z",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_fw <- logsplit.zph$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)


# FWdc_z:logtmid
plot(
  logsplit.zph,
  var = "FWdc_z:logtmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_log_fw <- logsplit.zph$table["FWdc_z:logtmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_log_fw, digits = 3)),
  cex = 1.1
)


dev.off()
```

Given that the log model appears to better fit the diagnostics, and that otherwise the models are basically equivalent by AIC, we will proceed with using the log model for wingsize time-dependence. 
### Sex-time interaction This diagnostic result points to a time-varying component of Sex, which we can aim to capture with an additional interaction term between Sex and time. We will fit this with log-transformed time as well.

```{r}
# Log-linear interaction
m_sex_tv_log <- coxph(
  Surv(tstart, tstop, event) ~
    Sex +
    FWdc_z +
    FWdc_z:logtmid +
    Sex:logtmid,
  data = ds
)

summary(m_sex_tv_log)
```

We can test the improvement in models via anova().

```{r}
anova_nine <- anova(m_log_split, m_sex_tv_log, test="LRT")
save_file(anova_nine,"Anova Comparison CoxPH model(Sex + Forewing discal cell + Forewing discal cell:logAge) vs. CoxPH model(Sex + Forewing discal cell + Forewing discal cell:logAge + Sex:logAge)", anova_stats)
```

The model comparison shows the log-time interaction is a significantly better fit. So the Sex:LogTime term significantly improves the model.

## Sex-time interactions

The diagnostics are not violated

```{r}
sextv.zph <- cox.zph(m_sex_tv_log)
sextv.zph
save_file(sextv.zph,"Diagnostics of CoxPH model(Forewing discal cell + Sex + Forewing discal cell:logAge + Size:logAge)", diagnostics_stats)
```

```{r log_time_diagnostics, fig.width=8, fig.height=8, dpi=300, out.width="180mm"}
par(mfrow = c(2,2))
yLim <- c(-5, 5)

# Sex
plot(
  sextv.zph,
  var = "Sex",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_sex <- sextv.zph$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)


# FWdc_z
plot(
  sextv.zph,
  var = "FWdc_z",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_fw <- sextv.zph$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)


# FWdc_z:logtmid
plot(
  sextv.zph,
  var = "FWdc_z:logtmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_log_fw <- sextv.zph$table["FWdc_z:logtmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_log_fw, digits = 3)),
  cex = 1.1
)

# Sex:logtmid
plot(
  sextv.zph,
  var = "Sex:logtmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_log_sex <- sextv.zph$table["Sex:logtmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_log_sex, digits = 3)),
  cex = 1.1
)
```


```{r save_log_time_plot}
jpeg("../Results/Survival_CoxPH/Figures/Figure_10_log_time_coxph_diagnostics.jpg",
     width = 180, height = 140, units = "mm", res = 300)

par(mfrow = c(2,2))
yLim <- c(-5, 5)

# Sex
plot(
  sextv.zph,
  var = "Sex",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_sex <- sextv.zph$table["Sex", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_sex, digits = 3)),
  cex = 1.1
)


# FWdc_z
plot(
  sextv.zph,
  var = "FWdc_z",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_fw <- sextv.zph$table["FWdc_z", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_fw, digits = 3)),
  cex = 1.1
)


# FWdc_z:logtmid
plot(
  sextv.zph,
  var = "FWdc_z:logtmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_log_fw <- sextv.zph$table["FWdc_z:logtmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_log_fw, digits = 3)),
  cex = 1.1
)

# Sex:logtmid
plot(
  sextv.zph,
  var = "Sex:logtmid",
  pch = 19,
  cex = 0.2,
  col = "red",
  ylim = yLim
)

p_log_sex <- sextv.zph$table["Sex:logtmid", "p"]

text(
  x = 0.5,
  y = 4.5,
  labels = paste0("p = ", format.pval(p_log_sex, digits = 3)),
  cex = 1.1
)

dev.off()
```

### Sex-time AIC

Let’s compute AIC on each of these models, even though the only ones we need this for are linear vs log.

```{r}
aic <- AIC(m_nosex_split, m_sex_split, 
          m_nosex_linear_split, m_linear_split, 
          m_nosex_log_split, m_log_split,
          m_sex_tv_log)

kbl(aic) %>%  kable_classic("striped") %>% 
  scroll_box(width = "60%")
```

### Interpretations

We will interpret the coefficients of our best model (log-time interaction). These are in “log hazard” scale: negative numbers mean a reduced hazard rate relative to the reference.

```{r}
coef(m_sex_tv_log)
```

Accordingly, for sex:

At early times, female hazard is lower than males; females are less likely to die. As time advances the HR approaches and then passes 1, and later in life, females hazard exceeds males. Here is a function, based on the model coefficients, to calculate the hazard ratio of females relative to males at any given time. Then we can calculate the M:F hazard ratio at a couple of time points.

```{r}
femHR <- function(timepoint) {
  coefs <- coef(m_sex_tv_log)
  beta_t <- coefs["SexFemale"] + coefs["SexFemale:logtmid"]*log(timepoint)
  hr_t <- exp(beta_t)
  return(hr_t)
}

times <- seq(5, 25, 5)
round(femHR(times), digits =  2)
```

Assuming an 'average-sized' butterfly, the hazard ratio bypasses 1 (meaning the relative impact of sex flips) at day 17.

```{r}
times <- c(1:40)
df <- data.frame(
  age = times,
  hazard_ratio = femHR(times)
)

hazard_ratio_plot <- ggplot(df, aes(age, hazard_ratio)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = 2, color = "red") +
  labs(
    x = "Age",
    y = "Female / Male Hazard Ratio"
  ) +
  theme_bw()
print(hazard_ratio_plot)
```

#### Save 'hazard ratio' plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_11_hazard_ratio.jpg",
  plot = hazard_ratio_plot,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

Size follows a similar pattern, but since it is a continuous variable it is slightly different to think about. In this case, since we normalized to Z-scale, our unit is standard deviation.

Accordingly, a larger size will have reduced hazard, at least early in life. However, that effect reduces as time increases. So later in life, size offers relatively little protection, and possibly hurts.

## Time transformation

With a continuous predictor like size, the survival package recommends using a “time transformation” (tt) function, which will appropriately capture the time-variation of such predictors via a mathematical function. This handles the “risk-set” issue in a dynamic way, and avoids the “borrowing information from the future” problem of fitting models with fixed-ages as predictors.

This is addressed in further detail in section 4.2 of the Time dependent analysis vignette from the survival package.

We can again fit all three versions of interactions between size and time. The problem here, as will see, is that this doesn’t allows us to examine the diagnostics with cox.zph(). Apparently, this diagnostic functiondoesn’t support models fit using tt().

Nonetheless, we’ll do this: 
### tt() linear Linear without Sex.

```{r}
m_nosex_tt_lin <- coxph(
  Surv(Age, event) ~ FWdc_z + tt(FWdc_z),
  data = monsurv,
  tt = function(x, t, ...) x * t
)
m_nosex_tt_lin_sum <- summary(m_nosex_tt_lin)
save_file(m_nosex_tt_lin_sum,"CoxPH Model(Forewing discal cell + Forewing discal cell:tt(Age))", coxph_stats)
print(m_nosex_tt_lin_sum)
```

```{r}
m_tt_lin <- coxph(
  Surv(Age, event) ~ Sex + FWdc_z + tt(FWdc_z),
  data = monsurv,
  tt = function(x, t, ...) x * t
)
m_tt_lin_sum <- summary(m_tt_lin)
save_file(m_tt_lin_sum,"CoxPH Model(Sex + Forewing discal cell + Forewing discal cell:tt(Age))", coxph_stats)
print(m_tt_lin_sum)
```

### tt() Logarithmic fit

Log without Sex

```{r}
m_nosex_tt_log <- coxph(
  Surv(Age, event) ~ FWdc_z + tt(FWdc_z),
  data = monsurv,
  tt = function(x, t, ...) x * log(t)
)
m_nosex_tt_log_sum <- summary(m_nosex_tt_log)
save_file(m_nosex_tt_log_sum,"CoxPH Model(Forewing discal cell + Forewing discal cell:log(tt(Age)))", coxph_stats)
print(m_nosex_tt_log_sum)
```

Log with Sex

```{r}
m_tt_log <- coxph(
  Surv(Age, event) ~ Sex + FWdc_z + tt(FWdc_z),
  data = monsurv,
  tt = function(x, t, ...) x * log(t)
)
m_tt_log_sum <- summary(m_tt_log)
save_file(m_tt_log_sum,"CoxPH Model(Sex + Forewing discal cell + Forewing discal cell:log(tt(Age)))", coxph_stats)
print(m_tt_log_sum)
```
```{r}
monsurv$Male <- as.numeric(monsurv$Sex == "Male")
m_tt_sex_log <- coxph(
  Surv(Age, event) ~ Sex + FWdc_z + tt(FWdc_z) + tt(Male),
  data = monsurv,
  tt = function(x, t, ...) x * log(t)
)
m_tt_sex_log_sum <- summary(m_tt_sex_log)
save_file(m_tt_log_sum,"CoxPH Model(Sex + Forewing discal cell + Forewing discal cell:log(tt(Age)) + Sex:log(tt(Age)))", coxph_stats)
print(m_tt_sex_log_sum)
```


### tt() model comparsions

The time varying term is already highly significant in the model fit itself. But we can’t do model comparisons directly via a typical LRT with the tt() (not supported in the package). However, we can still show a major difference in AIC, strongly supporting the better fit of the model with time-dependence in it.

Sex is not significant in the linear model comparison.

```{r}
anova_ten <- anova(m_nosex_tt_lin, m_tt_lin, test="LRT")
save_file(anova_ten,"Anova Comparison CoxPH Model(Forewing discal cell + Forewing discal cell:tt(Age)) vs. CoxPH Model(Sex + Forewing discal cell + Forewing discal cell:tt(Age))", anova_stats)
print(anova_ten)
```

```{r}
anova_eleven <- anova(m_nosex_tt_log, m_tt_log, test="LRT")
save_file(anova_eleven,"Anova Comparison CoxPH Model(Forewing discal cell + Forewing discal cell:log(tt(Age))) vs. CoxPH Model(Sex + Forewing discal cell + Forewing discal cell:log(tt(Age)))", anova_stats)
print(anova_eleven)
```

Sex is not significant in the logarithmic model comparison.

```{r}
anova_twelve <- anova(m_tt_log, m_tt_sex_log, test="LRT")
save_file(anova_twelve,"Anova Comparison CoxPH Model(Forewing discal cell + Forewing discal cell:log(tt(Age))) vs. CoxPH Model(Sex + Forewing discal cell + Forewing discal cell:log(tt(Age)) + Size:log(tt(Age)))", anova_stats)
print(anova_twelve)
```


## Compare models via AIC

We can’t compare between linear versus log with LRT because they have the same number of parameters are are not nested. Though we can examine AIC.

```{r}
aic <- AIC(m0_nosex, m0_sex, 
           m_nosex_tt_lin, m_tt_lin, 
           m_nosex_tt_log, m_tt_log, m_tt_sex_log)

kbl(aic) %>%  kable_classic("striped") %>% 
  scroll_box(width = "60%")
```

First, AIC clearly shows the tt() models with the effect of size time-varying are far better models. So even without the LRT comparison, this is strong support for including time-variation or size in the model.

As for linear vs log, parsimony, and the slightly better AIC, would argue for linear. But ultimately, it doesn’t really seem to matter, because it doesn’t substantively change the results of parameter estimates or significance. These models are arguably equivalent.

### Can’t diagnose tt()

Unfortunately, as mentioned above, Schoenfeld diagnostics does not function for time-transformation models.So we don’t really know how well the models are doing relative to assumptions 
### Model comparisons Here we are generating a table of results that report relevant models, comparisons via LRT, and the deltaAIC relative to the lowest model.

```{r}
# --------------------------------------------------
# Model comparison table for split-time CoxPH models
# --------------------------------------------------

model_fits <- list(
  M1 = m_nosex_split,
  M2 = m_sex_split,
  M3 = m_nosex_linear_split,
  M4 = m_linear_split,
  M5 = m_nosex_log_split,
  M6 = m_log_split,
  M7 = m_sex_tv_log
)

predictor_labels <- tibble(
  Model = names(model_fits),
  Predictors = c(
    "Size",
    "Size + Sex",
    "Size + Size × Time",
    "Size + Size × Time + Sex",
    "Size + Size × log(Time)",
    "Size + Size × log(Time) + Sex",
    "Size + Size × log(Time) + Sex + Sex × log(Time)"
  )
)

aic_tbl <- AIC(
  m_nosex_split,
  m_sex_split,
  m_nosex_linear_split,
  m_linear_split,
  m_nosex_log_split,
  m_log_split,
  m_sex_tv_log
) %>%
  as.data.frame() %>%
  rownames_to_column("fit_name") %>%
  mutate(
    Model = names(model_fits),
    delta_AIC = AIC - min(AIC)
  ) %>%
  select(Model, df, AIC, delta_AIC)

get_lrt <- function(fit_less, fit_more) {
  out <- anova(fit_less, fit_more, test = "LRT")
  
  tibble(
    LRT = out$Chisq[2],
    LRT_df = out$Df[2],
    p_value = out$`Pr(>|Chi|)`[2]
  )
}

lrt_tbl <- bind_rows(
  tibble(Model = "M1", Comparison = NA_character_, LRT = NA_real_, LRT_df = NA_real_, p_value = NA_real_),
  get_lrt(m_nosex_split, m_sex_split) %>% mutate(Model = "M2", Comparison = "M1 vs M2: + Sex"),
  get_lrt(m_nosex_split, m_nosex_linear_split) %>% mutate(Model = "M3", Comparison = "M1 vs M3: + Size × Time"),
  get_lrt(m_nosex_linear_split, m_linear_split) %>% mutate(Model = "M4", Comparison = "M3 vs M4: + Sex"),
  get_lrt(m_nosex_split, m_nosex_log_split) %>% mutate(Model = "M5", Comparison = "M1 vs M5: + Size × log(Time)"),
  get_lrt(m_nosex_log_split, m_log_split) %>% mutate(Model = "M6", Comparison = "M5 vs M6: + Sex"),
  get_lrt(m_log_split, m_sex_tv_log) %>% mutate(Model = "M7", Comparison = "M6 vs M7: + Sex × log(Time)")
) %>%
  select(Model, Comparison, LRT, LRT_df, p_value)

model_test_tbl <- predictor_labels %>%
  left_join(aic_tbl, by = "Model") %>%
  left_join(lrt_tbl, by = "Model") %>%
  mutate(
    across(c(AIC, delta_AIC, LRT, p_value), ~ round(.x, 4))
  )


write_csv(model_test_tbl, "../Results/Survival_CoxPH/Statistics/cox_split_time_model_comparison.csv")


kbl(model_test_tbl) %>%  kable_classic("striped") %>% 
  scroll_box(width = "100%")
```

```{r}
## Panel A: Female/Male hazard ratio
hazard_ratio_plot <- ggplot(df, aes(age, hazard_ratio)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = 2, color = "red") +
  labs(
    x = "Age",
    y = "Female / Male Hazard Ratio"
  ) +
  theme_bw() + ggtitle('A')
## Panel B: qx
qxLogplot <- ggplot(mon_sum_DF, aes(x = x, y = log(qx), color = strata)) +
  geom_line() +
  geom_point(size = 1) +
  scale_color_manual(values = sex_colors) +
  labs(x = "Age",
       y = "Log(Mortality)") +
  xlim(0, 45) +
  theme_bw() +
  theme(legend.position = "none") + ggtitle('B')

## Panel C: Smoothed log hazard
haz_df <- rbind(
  data.frame(
    age = haz_female$est.grid,
    loghaz = log(haz_female$haz.est),
    Sex = "Female"
  ),
  data.frame(
    age = haz_male$est.grid,
    loghaz = log(haz_male$haz.est),
    Sex = "Male"
  )
)

log_muhaz <- ggplot(haz_df, aes(x = age, y = loghaz, color = Sex)) +
  geom_line() +
  scale_color_manual(values = sex_colors) +
  labs(
    x = "Age",
    y = "Log(Mortality)"
  ) +
  coord_cartesian(ylim = c(-8, 0)) +
  theme_bw() +
  theme(legend.position = "none") + ggtitle('C')


final_fig <-
  hazard_ratio_plot | qxLogplot | log_muhaz +
  plot_annotation(tag_levels = "A")

final_fig
```

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_12_Figure3Panels.pdf",
  plot = final_fig,
  width = 280,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

```{r}
sessionInfo()
```
