---
title: "CoxPH"
author: Calvin Belton & James Walters
format: html
editor: source
theme: sandstone
toc: true
embed-resources: true
number-sections: true
---

## Setting up

```{r}
library(flexsurv)
library(ggplot2)
library(tidyverse)
library(ggfortify)# for `autoplot` with survfit
library(survminer)
library(kableExtra)
library(patchwork)
library(splines)
library(muhaz)
library(reporter)
```

# Load previously 'cleaned' dataset

```{r}
load(file = "../Input_Data/monsurv.Rdata")
```

## Function for saving files
```{r}
save_file <- function(test_obj, label, file){

  cat(
    "\n-----------------------------------\n",
    label,
    "\n-----------------------------------\n",
    file = file,
    append = TRUE
  )

  capture.output(
    test_obj,
    file = file,
    append = TRUE
  )
}
```

## Graphics
Before I start creating plots, I set up the custom theme and custom color palette that is continuous in my plots of monarch.
## Custom theme
```{r}
theme_monarch <- theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(face = "bold", size = 13),
    axis.text = element_text(face = "bold", size = 11),
    strip.text = element_text(face = "bold")
  )
```

### Graph systematics

```{r}
sex_colors <- c("Male" = "dodgerblue","Female" = "firebrick2")
colM <- "dodgerblue"
colF <- "firebrick2"
```

```{r}
survival_stats <- "../Results/Survival_CoxPH/Statistics/survival_stats.txt"

cat(
  "Monarch Manuscript Survival Results\n",
  "=================================\n\n",
  file = survival_stats
)
```

## Sex Comparison of Survival

## Create Survival Object

```{r}
survMon <- Surv(time = monsurv$Age, event = monsurv$event)
```

## Survdiff created: Kaplan-Mier statistics

```{r}
survMon.diff <- survdiff(survMon ~ Sex, data = monsurv)
print(survMon.diff)
save_file(survMon.diff,"Kaplan-Mier statistics", survival_stats)
```
#### Find the p-value

```{r}
p <- survMon.diff$pvalue
save_file(p,"Kaplan-Mier p-value", survival_stats)
```
### Analyze with survfit

```{r}
monsurv.fit <- survfit(Surv(Age) ~ Sex, data = monsurv)
print(monsurv.fit)
save_file(monsurv.fit,"Kaplan-Mier survfit survival estimation", survival_stats)
```
## Kaplan-Meier Plot

```{r}
survivorship_plot <- ggsurvplot(
  monsurv.fit,
  conf.int = FALSE,
  palette = sex_colors,
  xlab = "Days Since Eclosion",
  ylab = "Survival Probability",
  legend.title = "Sex",
  legend.labs = c("Male", "Female"),
  legend = "right",
  size = 1,
  ggtheme = theme_monarch
)
print(survivorship_plot)
```

#### Save Kaplan-Meier Plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_01_survivorship_plot.jpg",
  plot = survivorship_plot$plot,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

# Mortality rate visualization

### Calculate qx ()

##### Summarize previous survfit

```{r}
mon_sum <- summary(monsurv.fit)
str(mon_sum)
```
```{r}
mortality_stats <- "../Results/Survival_CoxPH/Statistics/mortality_statistics.txt"

cat(
  "Monarch Manuscript Mortality Results\n",
  "=================================\n\n",
  file = mortality_stats
)
```


```{r}
mon_sum_DF <- data.frame(
  x   = mon_sum$time,
  nx = mon_sum$n.risk,
  lx   = mon_sum$surv,
  dx = mon_sum$n.event,
  lower  = mon_sum$lower,
  upper  = mon_sum$upper,
  strata = sub(x = mon_sum$strata, pattern = "Sex=", replacement = "")
)
mon_sum_DF$qx <- mon_sum_DF$dx/mon_sum_DF$nx
save_file(mon_sum_DF,"Life-table data", mortality_stats)
```

### Visualize qx and log(qx) by sex

```{r}
qxplot <- ggplot(data = mon_sum_DF, aes(x=x, y = qx, col = strata)) +
  geom_line() +
  geom_point(size = 1) +
  scale_color_manual(values = c("firebrick2","dodgerblue"))+
  labs(title = "Mortality over time by Sex",
       x = NULL) +
  theme_bw()

qxLogplot <- ggplot(data = mon_sum_DF, aes(x=x, y =log(qx), col = strata)) +
  geom_line() +
  geom_point(size = 1) +
  labs(x = "Days" ) +
  scale_color_manual(values = c("firebrick2","dodgerblue"))+
  theme_bw()

# plot using patchwork syntax
mortality_plot <- qxplot / qxLogplot
print(mortality_plot)
```

#### Save mortality plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_02_mortality_plot.jpg",
  plot = mortality_plot,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

### Generate smoothed hazard rates via muhaz by sex

```{r}
# split by sex
haz_female <- muhaz(
  times = monsurv$Age[monsurv$Sex == "Female"],
  delta = monsurv$event[monsurv$Sex == "Female"],
  max.time = 45
)

haz_male <- muhaz(
  times = monsurv$Age[monsurv$Sex == "Male"],
  delta = monsurv$event[monsurv$Sex == "Male"],
  max.time = 45
)
```

### Overlay qx with smoothed hazard rates

```{r}
female_haz <- data.frame(
  time = haz_female$est.grid,
  hazard = log(haz_female$haz.est),
  sex = "Female"
)

male_haz <- data.frame(
  time = haz_male$est.grid,
  hazard = log(haz_male$haz.est),
  sex = "Male"
)

haz_df <- bind_rows(female_haz, male_haz)
obs_df <- mon_sum_DF %>%
  mutate(
    log_qx = log(qx),
    sex = strata
  )
ggplot() +
  geom_line(data = haz_df,
            aes(x = time, y = hazard, color = sex),
            linewidth = 1) +
  
  geom_point(data = obs_df,
             aes(x = x, y = log_qx, color = sex),
             size = 1.5) +
  
  facet_wrap(~sex, nrow = 1) +
  
  scale_color_manual(values = c("Female" = colF,
                                "Male" = colM)) +
  
  coord_cartesian(xlim = c(0, 50), ylim = c(-8, 0)) +
  
  labs(
    x = "Days",
    y = "Log(Hazard Rate)"
  ) +
  
  theme_classic() +
  theme(
    legend.position = "none"
  )
```
Save plot
```{r}
ggsave(
  "../Results/Survival_CoxPH/Figures/Figure_03_smoothed_hazard_plots.jpg",
  width = 180,
  height = 120,
  units = "mm",
  dpi = 300
)
```


# CoxPH with original wing length
```{r}
coxph_stats <- "../Results/Survival_CoxPH/Statistics/coxph_statistics.txt"

cat(
  "Monarch Manuscript Cox Proportional Hazards Results\n",
  "=================================\n\n",
  file = coxph_stats
)
```

### Model with winglength

```{r}
m0_nosex <- coxph(Surv(Age, event) ~ WingLength, data = monsurv)
sum_m0_nosex <- summary(m0_nosex)
save_file(sum_m0_nosex,"CoxPH model(WingLength)", coxph_stats)
```

### Model with winglength and sex

```{r}
m0_sex <- coxph(Surv(Age, event) ~ Sex + WingLength, data = monsurv)
sum_m0_sex <- summary(m0_sex)
save_file(sum_m0_sex,"CoxPH model(WingLength + Sex)", coxph_stats)
```

```{r}
anova_stats <- "../Results/Survival_CoxPH/Statistics/ANOVA_statistics.txt"

cat(
  "Monarch Manuscript Cox Proportional Hazards ANOVA Results\n",
  "=================================\n\n",
  file = anova_stats
)
```

### ANOVA comparison of the two previous models

```{r}
anova_one <- anova(m0_nosex, m0_sex, test="LRT")
save_file(anova_one,"Anova Comparison CoxPH model(WingLength) vs. CoxPH model(WingLength + Sex)", anova_stats)
```

## Contrast survival between males WITH winglength vs MISSING winglength

### Plot KM for two groups males

```{r}
monsurv$male_group <- NA

monsurv$male_group[monsurv$Sex == "Male" & is.na(monsurv$WingLength)] <- "NoWing"
monsurv$male_group[monsurv$Sex == "Male" & !is.na(monsurv$WingLength)] <- "Wing"

male_data <- subset(monsurv, Sex == "Male")

fit <- survfit(Surv(Age, event) ~ male_group, data = male_data)

male_plot <- ggsurvplot(
  fit, data = male_data,
  pval = FALSE,
  risk.table = FALSE,
  legend.title = "Male Group",
  legend.labs = c("No Winglength Measured", "Winglength Measured"),
  legend = c(0.8, 0.5),
  palette = c("forestgreen", "darkorange"),
  ggtheme = theme_monarch,
  xlab = "Days Since Eclosion"
)
print(male_plot)
```

#### Save male_plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_04_male_plot.jpg",
  plot = male_plot$plot,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

### Do KM test via survdiff for two groups of males (demonstrates that males missing wings are longer lived...)

```{r}
km_test <- survdiff(Surv(Age, event) ~ male_group, data = male_data)
print(km_test)
save_file(km_test,"Kaplan-Mier statistics for males data", survival_stats)
```

# CoxPH with Forward Discal Cell

```{r}
cuts <- seq(0, max(monsurv$Age), by = 1)  # 1-day intervals

ds <- survSplit(
  Surv(Age, event) ~ .,
  data = monsurv,
  cut = cuts[-c(1, length(cuts))],  
  start = "tstart",
  end = "tstop",
  event = "event"
)
# define analysis time for the interval 
ds$tmid <- (ds$tstart + ds$tstop) / 2
ds$logtmid <- log(ds$tmid)
```

### Fit split models

```{r}
m_nosex_split <- coxph(Surv(tstart, tstop, event) ~ FWdc_z, data = ds)

m_sex_split <- coxph(Surv(tstart, tstop, event) ~ Sex + FWdc_z, data = ds)

m_linear_split <-  coxph(Surv(tstart, tstop, event) ~ Sex + FWdc_z + FWdc_z:tmid, data = ds)

m_log_split <- coxph(Surv(tstart, tstop, event) ~ Sex + FWdc_z + FWdc_z:logtmid, data = ds)

m_spline_split <- coxph(Surv(tstart, tstop, event) ~ Sex + FWdc_z + FWdc_z:ns(tmid, df=3), data = ds)
```

```{r}
m_nosex_split_sum <- summary(m_nosex_split)
save_file(m_nosex_split_sum,"CoxPH Model(Forewing discal cell)", coxph_stats)
```

```{r}
m_sex_split_sum <- summary(m_sex_split)
save_file(m_sex_split_sum,"CoxPH Model(Forewing discal cell + Sex)", coxph_stats)
```

```{r}
m_linear_split_sum <- summary(m_linear_split)
save_file(m_linear_split_sum,"CoxPH Model(Forewing discal cell + Sex + Linear time interaction with Forewing discal cell)", coxph_stats)
```

```{r}
m_log_split_sum <- summary(m_log_split)
save_file(m_log_split_sum,"CoxPH Model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell)", coxph_stats)
```

```{r}
m_spline_split_sum <- summary(m_spline_split)
save_file(m_spline_split_sum,"CoxPH Model(Forewing discal cell + Sex + Spline time interaction with Forewing discal cell)", coxph_stats)
```


### Split model tests

Comparisons among models indicates that the spline model is not a significantly better fit. Thus, we proceed with the log-linear model.

```{r}
anova_two <- anova(m_nosex_split, m_sex_split, test = "LRT")
save_file(anova_two,"Anova Comparison CoxPH model(Forewing discal cell) vs. CoxPH model(Forewing discal cell + Sex)", anova_stats)
```

```{r}
anova_three <- anova(m_sex_split, m_linear_split, test = "LRT")
save_file(anova_three,"Anova Comparison CoxPH model(Forewing discal cell + Sex) vs. CoxPH model(Forewing discal cell + Sex + Linear time interaction with Forewing discal cell)", anova_stats)
```

```{r}
anova_four <- anova(m_linear_split, m_log_split, test = "LRT")
save_file(anova_four,"Anova Comparison CoxPH model(Forewing discal cell + Sex + Linear time interaction with Forewing discal cell) vs. CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell)", anova_stats)
```

```{r}
anova_five <- anova(m_log_split, m_spline_split, test = "LRT")
save_file(anova_five,"Anova Comparison CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell) vs. CoxPH model(Forewing discal cell + Sex + Spline time interaction with Forewing discal cell)", anova_stats)
```

### Split diagnostics log-linear
```{r}
diagnostics_stats <- "../Results/Survival_CoxPH/Statistics/diagnostics_statistics.txt"

cat(
  "Monarch Manuscript Diagnostics of Cox Proportional Hazards Models\n",
  "=================================\n\n",
  file = diagnostics_stats
)
```

We can examine the diagnostics for this model fit now. What we find in this case is that Sex now violates the diagnostics, seemingly invoking the mid-life shifts we observed in the plots of interval hazard rates.

```{r}
logsplit.zph <- cox.zph(m_log_split)
print(logsplit.zph)
save_file(logsplit.zph,"Diagnostics of CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell)", diagnostics_stats)
```
View plot in markdown
```{r split_coxph_diagnostics, fig.width=8, fig.height=8, dpi=300, out.width="180mm"}
par(mfrow = c(2,2))
plot(logsplit.zph)
```
Save plot
```{r split_coxph_plot, eval=FALSE}
jpeg(
  filename = "../Results/Survival_CoxPH/Figures/Figure_05_split_coxph_diagnostics.jpg",
  width = 180,
  height = 120,
  units = "mm",
  res = 300
)

par(mfrow = c(2,2))
plot(logsplit.zph)

dev.off()
```
### Sex-time interaction

This diagnostic result points to a time-varying component of Sex, which we can aim to capture with an additional interaction term between Sex and time

```{r}
# Linear interaction
m_sex_tv_lin <- coxph(
  Surv(tstart, tstop, event) ~
    Sex +
    FWdc_z +
    FWdc_z:logtmid +
    Sex:tmid,
  data = ds
)
m_sex_tv_lin_sum <- summary(m_sex_tv_lin)
save_file(m_sex_tv_lin_sum,"CoxPH Model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell + Linear time interaction with Sex)", coxph_stats)
```

```{r}
# Log-linear interaction
m_sex_tv_log <- coxph(
  Surv(tstart, tstop, event) ~
    Sex +
    FWdc_z +
    FWdc_z:logtmid +
    Sex:logtmid,
  data = ds
)
m_sex_tv_log_sum <- summary(m_sex_tv_log)
save_file(m_sex_tv_log_sum,"CoxPH Model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell + Logarithmic time interaction with Sex)", coxph_stats)
```

```{r}
anova_six <- anova(m_log_split, m_sex_tv_lin, test="LRT")
print(anova_six)
save_file(anova_six,"Anova Comparison CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell) vs. CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell + Linear time interaction with Sex)", anova_stats)
```

```{r}
anova_seven <- anova(m_sex_tv_lin, m_sex_tv_log, test="LRT")
print(anova_seven)
save_file(anova_seven,"Anova Comparison CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell + Linear time interaction with Sex) vs. CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell + Logarithmic time interaction with Sex)", anova_stats)
```

The model comparison show that the log-time interaction is a better fit than the linear-time interaction. The Sex:LogTime term significantly improves the model.

The diagnostics are not violated

```{r}
sextv.zph <- cox.zph(m_sex_tv_log)
sextv.zph
save_file(sextv.zph,"Diagnostics of CoxPH model(Forewing discal cell + Sex + Logarithmic time interaction with Forewing discal cell + Logarithmic time interaction with Sex)", diagnostics_stats)
```

```{r log_time_diagnostics, fig.width=8, fig.height=8, dpi=300, out.width="180mm"}
par(mfrow = c(2,2))
plot(sextv.zph)
```

```{r save_log_time_plot, eval=FALSE}
jpeg("../Results/Survival_CoxPH/Figures/Figure_06_log_time_coxph_diagnostics.jpg",
     width = 180, height = 120, units = "mm", res = 300)

par(mfrow = c(2,2))
plot(sextv.zph)

dev.off()
```

### Interpretations

We will interpret the coefficients of our best model (log-time interaction). These are in “log hazard” scale: negative numbers mean a reduced hazard rate relative to the reference.

```{r}
coef(m_sex_tv_log)
```

Interpreting this, for sex:

At early times, female hazard is lower than males; females are less likely to die. As time advances the hazard ratio approaches and then passes 1, and later in life, females hazard exceeds males.

Here is a function, based on the model coefficients, to calculate the hazard ratio of females relative to males at any given time. Then we can calculate the M:F hazard ratio at a couple of time points.

```{r}
femHR <- function(timepoint) {
  coefs <- coef(m_sex_tv_log)
  beta_t <- coefs["SexFemale"] + coefs["SexFemale:logtmid"]*log(timepoint)
  hr_t <- exp(beta_t)
  return(hr_t)
}

times <- seq(5, 25, 5)
round(femHR(times), digits =  2)
```

Assuming an 'average-sized' butterfly, the hazard ratio bypasses 1 (meaning the relative impact of sex flips) at day 17.

```{r}
times <- c(1:40)
hazard_ratio <- plot(x = times, 
     y = femHR(times),
     type="l",
     ylab="Female / Male Hazard Ratio",
     xlab="Age")
abline(h=1, lty=2, col = "red")
print(hazard_ratio)
```

#### Save 'hazard ratio' plot

```{r}
ggsave(
  filename = "../Results/Survival_CoxPH/Figures/Figure_07_hazard_ratio.jpg",
  plot = hazard_ratio,
  width = 180,    
  height = 120,
  units = "mm",
  dpi = 300
)
```

Size follows a similar pattern, but since it is a continuous variable it is slightly different to think about. In this case, since we normalized to Z-scale, our unit is standard deviation.

Accordingly, a larger size will have reduced hazard, at least early in life. However, that effect reduces as time increases. So later in life, size offers relatively little protection, and possibly hurts.

```{r}
sessionInfo()
```
